---
title: Currency Exchange Rates with Bayesian
author: Priscilla Chen
format:
  html:
    code-tools: true
    toc: true
    code-fold: true
    embed-resources: true
    self-contained: true
---

## Main Question:

How do Differentials in Interest rates, Inflation, and Country's economical Growth impact the currency exchange rate between US dollars and Chinese Yuan?

## Data Preperation

```{r, warning=FALSE}
#| include: false
library(dagitty)
library(ggdag)
library(CalvinBayes)
library(dplyr)
# library(mosaic)
library(tidyverse)
library(readr)
library(rstan)
library(ggformula)
library(rethinking)
library(bayesplot)
library(tidybayes)
```

Data file 1: Monthly Exchange Rate between US dollars and Chinese Yuan, extracted from [Federal Researve](https://www.federalreserve.gov/releases/h10/current/) (Data range from 1990-2 to 2025-2)

```{r, echo=FALSE}
Exchange_rate <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/FRB_H10 (1).csv", skip = 5)
Exchange_rate <- Exchange_rate |>
  mutate(Time.Period = paste0(Time.Period, "-01"))|>
  rename(Exchange_rate = RXI_N.M.CH,
         observation_date = Time.Period)

Exchange_rate$observation_date <- as.Date(Exchange_rate$observation_date)
```

Data File 2: Inflation: YEARLY Consumer Prices for USA from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/FPCPITOTLZGUSA). (Note this dataset only contain data till 2023)

```{r,echo=FALSE}
Inflation_US <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/FPCPITOTLZGUSA.csv")
Inflation_US <- Inflation_US |>
  rename(Inflation_US = FPCPITOTLZGUSA )

# create a copy
inflation_data <- Inflation_US
inflation_data$observation_date <- as.Date(inflation_data$observation_date)
# extract year series
year_vector <- as.numeric(format(inflation_data$observation_date, "%Y"))
# extract yearly Inflation value
yearly_values <- inflation_data$Inflation_US
# creating a monthly series
monthly_dates <- seq(from = as.Date("1987-01-01"), to = as.Date("2023-01-01"), by = "month")

# convert using linear interpolation
monthly_year_fraction <- as.numeric(format(monthly_dates, "%Y")) +
                         (as.numeric(format(monthly_dates, "%m")) - 1) / 12
monthly_inflation <- approx(x = year_vector, y = yearly_values, xout = monthly_year_fraction)$y

# Combine data
monthly_inflation_US <- data.frame(observation_date = monthly_dates, Monthly_Inflation_US = monthly_inflation)
```

Data File 3: Inflation: YEARLY Consumer Prices for China from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/FPCPITOTLZGCHN). (Note this dataset only contain data till 2023)

```{r, echo=FALSE}
Inflation_CN <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/FPCPITOTLZGCHN.csv")
Inflation_CN <- Inflation_CN |>
  rename(Inflation_CN = FPCPITOTLZGCHN)
# create a copy
inflation_data <- Inflation_CN
inflation_data$observation_date <- as.Date(inflation_data$observation_date)
# extract year series
year_vector <- as.numeric(format(inflation_data$observation_date, "%Y"))
# extract yearly Inflation value
yearly_values <- inflation_data$Inflation_CN
# creating a monthly series
monthly_dates <- seq(from = as.Date("1987-01-01"), to = as.Date("2023-01-01"), by = "month")

# convert using linear interpolation
monthly_year_fraction <- as.numeric(format(monthly_dates, "%Y")) +
                         (as.numeric(format(monthly_dates, "%m")) - 1) / 12
monthly_inflation <- approx(x = year_vector, y = yearly_values, xout = monthly_year_fraction)$y

# Combine data
monthly_inflation_CN <- data.frame(observation_date = monthly_dates, Monthly_Inflation_CN = monthly_inflation)

```


Data File 4: Monthly GDP Ratio to Trend for USA, downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/USALORSGPRTSTSAM). (Note this dataset only contain data till 2023-11-01)

```{r, echo=FALSE}
GDP_US <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/USALORSGPNOSTSAM.csv")
GDP_US <- GDP_US |>
  rename(GDP_US = USALORSGPNOSTSAM )
GDP_US$observation_date <- as.Date(GDP_US$observation_date)
```

Data File 5: Monthly GDP Ratio to Trend for China, downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/CHNLORSGPRTSTSAM). (Note this dataset only contain data till 2024-02-01)

```{r, echo=FALSE}
GDP_CN <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/CHNLORSGPNOSTSAM.csv")
GDP_CN <- GDP_CN |>
  rename(GDP_CN = CHNLORSGPNOSTSAM)
GDP_CN$observation_date <- as.Date(GDP_CN$observation_date)
```

Data file 6: Monthly US Interest Rate (Rates which Federal Reserve charge on banks for overnight borrow), downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/INTDSRUSM193N). (Note this dataset only contain data till 2021-08-01)

```{r, echo=FALSE}
Interest_US <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/INTDSRUSM193N.csv")
Interest_US <- Interest_US |>
  rename(Interest_US = INTDSRUSM193N)

Interest_US$observation_date <- as.Date(Interest_US$observation_date)
```

Data file 7: Monthly Chinese Interest Rate (Rates which Federal Reserve charge on banks for overnight borrow), downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/INTDSRCNM193N)

```{r, echo=FALSE}
Interest_CN <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/INTDSRCNM193N.csv")
Interest_CN <- Interest_CN |>
  rename(Interest_CN = INTDSRCNM193N)
Interest_CN$observation_date <- as.Date(Interest_CN$observation_date)
```

**Joined Data:**

```{r}
datasets <- list(Exchange_rate,GDP_CN,GDP_US, monthly_inflation_CN, monthly_inflation_US, Interest_CN, Interest_US)

data <- reduce(datasets, full_join, by = "observation_date")

data <- data|>
  mutate(Dif_GDP = GDP_US - GDP_CN,
         Dif_Interest_Rate = Interest_US - Interest_CN,
         Dif_Inflation_Rate = Monthly_Inflation_US - Monthly_Inflation_CN)|>
  # "first difference"
  mutate(delta_exch_rate = Exchange_rate - lag(Exchange_rate))
# get rid of the first two row because there's some missing data. 
data <- data[-c(1, 2), ]
# glimpse(data)

# scaling all variables other than observation_date. 
scaled_data <- data %>%
  mutate(across(
    .cols = where(is.numeric),
    .fns = ~ scale(.) %>% as.vector
  ))

scaled_data <- scaled_data |>
  select(Dif_GDP, Dif_Interest_Rate, Dif_Inflation_Rate, delta_exch_rate)

data_with_na <- scaled_data

scaled_data <- na.omit(scaled_data)

glimpse(scaled_data)
```

## Casual Diagram:

[Economics Help](https://www.economicshelp.org/macroeconomics/exchangerate/factors-influencing/) listed 5 factors that would affect exchange rate: rate of growth, Interest rate, Inflation rate, Competitiveness, and Speculations. Since competitiveness and Speculations are hard to measure quantitatively, they are not included. Besides, rate of growth is measured by GDP ratio.


```{r}
causal_diagram <- dagitty("dag{
  Interest_Rate_dif -> delta_Exchange_Rate;
  Inflation_Rate_dif -> delta_Exchange_Rate;
  Interest_Rate_dif -> Inflation_Rate_dif;
  Growth_Rate_dif -> delta_Exchange_Rate;
}")
gg_dag(causal_diagram)
```

## Model Description

$$
\text{Exchange Rate Difference Scaled} \sim \text{Normal}(\mu_i, \sigma) 
$$
$$\mu_i \sim \beta_0 + \beta_1 * \text{Interest Rate Differential Scaled} + \beta_2 * \text{Inflation Rate Differential Scaled} + \beta_3 * \text{Growth Rate Differential Scaled} $$
$$
\beta_0 \sim \text{Normal}(0,1)
$$
$$
\beta_1 \sim \text{Normal}(0.5,1)
$$
$$
\beta_2 \sim \text{Normal}(-0.5,1)
$$
$$
\beta_3 \sim \text{Normal}(0.5,1)
$$
$$
\sigma \sim \text{Exponential}(1)
$$

## Rationals for Prior

Among the key determinants, differentials in interest rates, inflation rates, and gross domestic product growth between US and China have been widely studied as fundamental economic forces influencing currency exchange values. Prior research with Autoregressive distributed lag (ARDL) model indicates that gross domestic product growth have a positive effect on the exchange rate between USD and CNY, while interest rate and inflation rates have a negative effect (Khan, M.K.,2019). While most research agrees on the positive effect of GDP/economic growth and the negative effects of inflation rate toward exchange rate, findings on the relationship between interest rate and currency exchange rate is equivocal. Some studies suggest that an increase in US Federal Reserve's interest rate would lead to depreciation of CNY and appreciation in USD (Zhou and Zhang, Li). 

Since higher inflation rate may depreciate the currency and thus lowers the exchange rate, if Dif_inflation_rate increases, delta_dif_exchange_rate should decrease. I'm using a normal distribution center at -0.5 and has the standard deviation of 1. This means that for one standard deviation increase in differential in inflation rate, I suspect the differential in delta exchange rate will most likely decrease by 0.5 standard deviation. However, there's still possibility for a wide range of positive and negative slopes. 


Since there are ambivalent findings on the relationship between interest rate and exchange rate, but more studies lean toward a positive relationship, theorizing a higher interest rate attract foreign investers and drive up demand for the currency (Li). Therefore, if dif_interest_rate (US - CN) increases, delta_dif_exchange_rate will increase, meaning US currency will worth more in China currency. For its slope, beta 1, I'm using a normal distribution center at 0.5 and has the standard deviation of 1. This means that for one standard deviation increase in differential in interest rate, the delta differential in exchange rate will most likely increase by 0.5 standard deviation. However, there is still possibility for a wider range of slope from negative to positive.


GDP measures growth. The better the growth, the better the currency for it attracts foreign investers, driving up the needs for the currency(Khan, M.K.). Therefore, if Dif_GDP increase,  delta_dif_exchange_rate should also increase. For the slope for Dif_GDP, Beta 3, I'm using a normal distribution centering at 0.5 with standard deviation of 1. This prior suggests a likely possitive relationship between GDP and exchange rate, but still keep the possibility for a wide range of both positive and negative values.

I suspect that if all the above are kept at 0, there would be no change in exchange rate. Therefore I used a normal distribution that centers at 0 with standard deviation of 1 for Beta 0, the intercept. 

I'm using exponential(1) as the prior for sigma, with lower bound at 0, because the standard deviation cannot be 0. 

## Prior Predictive Distribution
```{r}
n_sim <- 500


predictors <- scaled_data |>
  select(Dif_Interest_Rate, Dif_Inflation_Rate, Dif_GDP) |>
  rename(interest = Dif_Interest_Rate,
         inflation = Dif_Inflation_Rate,
         growth = Dif_GDP) |>
  mutate(obs_id = row_number())

priors <- tibble(
  sim_id = 1:n_sim,
  beta_0 = rnorm(n_sim, 0, 1),
  beta_1 = rnorm(n_sim, 0.5, 1),
  beta_2 = rnorm(n_sim, -0.5, 1),
  beta_3 = rnorm(n_sim, 0.5, 1),
  sigma = rexp(n_sim, 1)
)

prior_pred <- crossing(priors, predictors) |>
  mutate(
    mu = beta_0 + beta_1 * interest + beta_2 * inflation + beta_3 * growth,
    y_sim = rnorm(n(), mean = mu, sd = sigma)
  )|>
  ungroup()

```
```{r}
gf_histogram(~y_sim,
             data = prior_pred)
```

The prior predictive distribution show the simulated delta_exchange_rate differences centers at 0, and have a range from about -12 to 12, which is large for a scaled z-score variable, but this wide range can make sure we don't exclude possible values. 



## Fitting Model

```{r,results='hide'}
stan_data <- compose_data(scaled_data)
# fit the model
exch_rate_model <- rstan::stan(file = 'model1.stan', data = stan_data)

```
```{r}
exch_rate_model
```

```{r}
bayesplot::mcmc_trace(exch_rate_model, pars = c("b0", "b1", "b2", "b3", "sigma"))
```

## Model Diagnosis

**n_eff: ** n_eff is effective number of samples drawn which is an estimate of the number of independent samples from the posterior distribution. The model above displays large n_eff numbers which indicate a relatively good model.

**Rhat: ** reached exactly 1 for the model above, meaning the chain converge and well-mixed, indicating a good model.

**Trace Plot: ** The trace plot demonstrate the following aspects, indicating the model above is good to draw conclusions from. 

  - Stationary: the path of each chain staying within the same high-probability portion of the posterior distribution. In the plot, the zigzag traces all stays relatively around a stable central tendency.

  - Good-Mixing: the zigzag show that the chain doesn’t slowly wander, but rather rapidly explores the full region.

  - Convergence: multiple, independent chains stick around the same region of high probability.

## Model Comparison

I fitted another model with less predictors:
```{r,results='hide'}
scaled_data2 <- scaled_data|>
  select(Dif_Inflation_Rate, Dif_GDP, delta_exch_rate)
stan_data2 <- compose_data(scaled_data2)
# model without Differential in Interest Rate predictor
exch_rate_model2 <- rstan::stan(file = 'model2.stan', data = stan_data2)
```

```{r,results='hide'}
# without Dif_Inflation_Rate
scaled_data3 <- scaled_data|>
  select(Dif_Interest_Rate, Dif_GDP, delta_exch_rate)
stan_data3 <- compose_data(scaled_data3)
# model without Differential in Inflation Rate predictor
exch_rate_model3 <- rstan::stan(file = 'model3.stan', data = stan_data3)
```
```{r,results='hide'}
# without GDP
scaled_data4 <- scaled_data|>
  select(Dif_Interest_Rate, Dif_Inflation_Rate, delta_exch_rate)
stan_data4 <- compose_data(scaled_data4)
# model without Differential in GDP predictor
exch_rate_model4 <- rstan::stan(file = 'model4.stan', data = stan_data4)
```



```{r}
# Model Comparison
rethinking::compare(exch_rate_model,exch_rate_model2,exch_rate_model3,exch_rate_model4, func = WAIC)
```
```{r}
# Model Comparison
rethinking::compare(exch_rate_model,exch_rate_model2,exch_rate_model3,exch_rate_model4, func = PSIS)
```
According to WAIC test, exch_rate_model4 with two predictor, differential in Interest Rate and differential in Inflation Rate, is better than other models as its WAIC score is the lowest among the different models. 
According to PSIS, however, exch_rate_model with all three predictors, differential in Interest Rate, differential in Inflation Rate, and differential in GDP, as its PSIS score is the lowest among the different models. 

## Posteriors

```{r}
posterior_samples <- as.data.frame(exch_rate_model)

posterior_long <- posterior_samples %>%
  pivot_longer(cols = c(b0, b1, b2, b3, sigma), names_to = "parameter", values_to = "value")

gf_dens(~value | parameter, data = posterior_long, group = ~parameter, color = ~parameter, size = 1)
```


### Interpret Posterior:

b0 centers around 0.01, suggesting that if predictors such as differentials in interest rate, differentials in inflation, and differentials in GDP are controlled as 0, the change in exchange rate would likely be 0.01, a very slight positive increase of the exchange rate between US dollars and Chinese Yuan. 

b1 centers around - 0.05, suggesting that 1 standard deviation increase in the differentials in interest rate (US - CN) is correlated with 0.05 standard deviation decrease in exchange rate. In other words, as US has a higher interest rate than Chinese Yuan, the value of US dollars will very slightly decrease in term of the Chinese Yuan currency. 

b2 centers around -0.2, suggesting that 1 standard deviation increase in the differentials in inflation (US - CN) is correlated with 0.2 decrease in exchange rate. In other words, as US dollars experience a higher inflation rate than Chinese Yuan, the value of US dollars will slightly decrease in terms of Chinese Yuan currency.

b3 centers around 0.08, suggesting that 1 standard deviation increase in the differentials in GDP (US - CN) is correlated with 0.08 increase in exchange rate. In other words, as US experience a higher economic growth than China, measured by GDP, the value of US dollars will slightly increase in terms of Chinese Yuan currency.

sigma centers around 1.01, suggesting that the standard deviation for the exchange rate is aorund 1.01. 

### Discussion & Conclusion:

The model result suggests that: 

(1) there's a slight negative relationship between differentials in interest rate and change in exchange rate, such that as US interest rate is higher than Chinese Yuan for 1 standardard deviation, the value of US dollars **decrease** 0.05 standard deviation in term of the Chinese Yuan currency. This result contradict other sources in that most articles suggest higher interest rate can increase exchange rate by attracting investors and increasing currency values

(2) there's a negative relationship between differentials in inflation rate and change in exchange rate, such that as US inflation rate is higher than Chinese Yuan for 1 standardard deviation, the value of US dollars **decrease** 0.2 standard deviation in term of the Chinese Yuan currency. This result align with other sources which suggest rising inflation rate can depreciate currency value and thus lead to lower exchange rate. 

(3) there's a positive relationship between differentials in country's economic growth and change in exchange rate, such that as US GDP is higher than Chinese GDP for 1 standard deviation, the value of US dollars **increase** 0.08 standard deviation in term of the Chinese Yuan currency. This finding align with other sources as well, which suggest rising economic growht increase currency value and attract investors, leading to higher exchange rate.

### Work Cited

Khan, M.K., Teng, JZ. & Khan, M.I. *Cointegration between macroeconomic factors and the exchange rate USD/CNY.* Financ Innov 5, 5 (2019). https://doi.org/10.1186/s40854-018-0117-x

Li, Jie. *The Impact of the Interest Rates Raised by Federal Reserve System on the Exchange Rate of the US Dollar Against the Chinese Yuan.* 2023, https://www.researchgate.net/publication/376131422.

Zhou, Hao, and Qiang Zhang. *Analysis of Interest Rate Hikes and Exchange Rate Between U.S. and China.* 2023, https://www.researchgate.net/publication/376132090.


