---
title: Currency Exchange Rates with Bayesian
author: Priscilla Chen
format:
  html:
    code-tools: true
    toc: true
    embed-resources: true
    self-contained: true
---

## Main Question:

How do Differentials in Interest rates, Inflation, and Country's economical Growth impact the currency exchange rate between US dollars and Chinese Yuan?

How confidently can we predict currency exchange rate from Differentials in Interest rates, Inflation, and Country's Economical Growth?

## Data Preperation

```{r, warning=FALSE}
#| include: false
library(dagitty)
library(ggdag)
library(CalvinBayes)
library(dplyr)
# library(mosaic)
library(tidyverse)
library(readr)
library(rstan)
library(ggformula)
library(rethinking)
library(bayesplot)
library(tidybayes)
```

Data file 1: Monthly Exchange Rate between US dollars and Chinese Yuan, extracted from [Federal Researve](https://www.federalreserve.gov/releases/h10/current/) (Data range from 1990-2 to 2025-2)

```{r, echo=FALSE}
Exchange_rate <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/FRB_H10 (1).csv", skip = 5)
Exchange_rate <- Exchange_rate |>
  mutate(Time.Period = paste0(Time.Period, "-01"))|>
  rename(Exchange_rate = RXI_N.M.CH,
         observation_date = Time.Period)

Exchange_rate$observation_date <- as.Date(Exchange_rate$observation_date)
```

Data File 2: Inflation: YEARLY Consumer Prices for USA from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/FPCPITOTLZGUSA). (Note this dataset only contain data till 2023)

```{r,echo=FALSE}
Inflation_US <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/FPCPITOTLZGUSA.csv")
Inflation_US <- Inflation_US |>
  rename(Inflation_US = FPCPITOTLZGUSA )

# create a copy
inflation_data <- Inflation_US
inflation_data$observation_date <- as.Date(inflation_data$observation_date)
# extract year series
year_vector <- as.numeric(format(inflation_data$observation_date, "%Y"))
# extract yearly Inflation value
yearly_values <- inflation_data$Inflation_US
# creating a monthly series
monthly_dates <- seq(from = as.Date("1987-01-01"), to = as.Date("2023-01-01"), by = "month")

# convert using linear interpolation
monthly_year_fraction <- as.numeric(format(monthly_dates, "%Y")) +
                         (as.numeric(format(monthly_dates, "%m")) - 1) / 12
monthly_inflation <- approx(x = year_vector, y = yearly_values, xout = monthly_year_fraction)$y

# Combine data
monthly_inflation_US <- data.frame(observation_date = monthly_dates, Monthly_Inflation_US = monthly_inflation)
```

Data File 3: Inflation: YEARLY Consumer Prices for China from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/FPCPITOTLZGCHN). (Note this dataset only contain data till 2023)

```{r, echo=FALSE}
Inflation_CN <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/FPCPITOTLZGCHN.csv")
Inflation_CN <- Inflation_CN |>
  rename(Inflation_CN = FPCPITOTLZGCHN)
# create a copy
inflation_data <- Inflation_CN
inflation_data$observation_date <- as.Date(inflation_data$observation_date)
# extract year series
year_vector <- as.numeric(format(inflation_data$observation_date, "%Y"))
# extract yearly Inflation value
yearly_values <- inflation_data$Inflation_CN
# creating a monthly series
monthly_dates <- seq(from = as.Date("1987-01-01"), to = as.Date("2023-01-01"), by = "month")

# convert using linear interpolation
monthly_year_fraction <- as.numeric(format(monthly_dates, "%Y")) +
                         (as.numeric(format(monthly_dates, "%m")) - 1) / 12
monthly_inflation <- approx(x = year_vector, y = yearly_values, xout = monthly_year_fraction)$y

# Combine data
monthly_inflation_CN <- data.frame(observation_date = monthly_dates, Monthly_Inflation_CN = monthly_inflation)

```


Data File 4: Monthly GDP Ratio to Trend for USA, downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/USALORSGPRTSTSAM). (Note this dataset only contain data till 2023-11-01)

```{r, echo=FALSE}
GDP_US <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/USALORSGPNOSTSAM.csv")
GDP_US <- GDP_US |>
  rename(GDP_US = USALORSGPNOSTSAM )
GDP_US$observation_date <- as.Date(GDP_US$observation_date)
```

Data File 5: Monthly GDP Ratio to Trend for China, downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/CHNLORSGPRTSTSAM). (Note this dataset only contain data till 2024-02-01)

```{r, echo=FALSE}
GDP_CN <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/CHNLORSGPNOSTSAM.csv")
GDP_CN <- GDP_CN |>
  rename(GDP_CN = CHNLORSGPNOSTSAM)
GDP_CN$observation_date <- as.Date(GDP_CN$observation_date)
```

Data file 6: Monthly US Interest Rate (Rates which Federal Reserve charge on banks for overnight borrow), downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/INTDSRUSM193N). (Note this dataset only contain data till 2021-08-01)

```{r, echo=FALSE}
Interest_US <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/INTDSRUSM193N.csv")
Interest_US <- Interest_US |>
  rename(Interest_US = INTDSRUSM193N)

Interest_US$observation_date <- as.Date(Interest_US$observation_date)
```

Data file 7: Monthly Chinese Interest Rate (Rates which Federal Reserve charge on banks for overnight borrow), downloaded from [Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/series/INTDSRCNM193N)

```{r, echo=FALSE}
Interest_CN <- read.csv("C:/Users/prisc/OneDrive - Calvin University/2Spring/Stat341/Currency-Exchange-Rate-Project/datafiles/INTDSRCNM193N.csv")
Interest_CN <- Interest_CN |>
  rename(Interest_CN = INTDSRCNM193N)
Interest_CN$observation_date <- as.Date(Interest_CN$observation_date)
```

**Joined Data:**

```{r}
datasets <- list(Exchange_rate,GDP_CN,GDP_US, monthly_inflation_CN, monthly_inflation_US, Interest_CN, Interest_US)

data <- reduce(datasets, full_join, by = "observation_date")

data <- data|>
  mutate(Dif_GDP = GDP_US - GDP_CN,
         Dif_Interest_Rate = Interest_US - Interest_CN,
         Dif_Inflation_Rate = Monthly_Inflation_US - Monthly_Inflation_CN)|>
  # "first difference"
  mutate(delta_exch_rate = Exchange_rate - lag(Exchange_rate))
# get rid of the first two row because there's some missing data. 
data <- data[-c(1, 2), ]
# glimpse(data)

# scaling all variables other than observation_date. 
scaled_data <- data %>%
  mutate(across(
    .cols = where(is.numeric),
    .fns = ~ scale(.) %>% as.vector
  ))
```
```{r}
scaled_data <- scaled_data |>
  select(Dif_GDP, Dif_Interest_Rate, Dif_Inflation_Rate, delta_exch_rate)
scaled_data <- na.omit(scaled_data)
glimpse(scaled_data)
```

## Casual Diagram:

[Economics Help](https://www.economicshelp.org/macroeconomics/exchangerate/factors-influencing/) listed 5 factors that would affect exchange rate: rate of growth, Interest rate, Inflation rate, Competitiveness, and Speculations. Since competitiveness and Speculations are hard to measure quantitatively, they are not included. Besides, rate of growth is measured by GDP ratio.


```{r}
causal_diagram <- dagitty("dag{
  Interest_Rate_dif -> delta_Exchange_Rate;
  Inflation_Rate_dif -> delta_Exchange_Rate;
  Interest_Rate_dif -> Inflation_Rate_dif;
  Growth_Rate_dif -> delta_Exchange_Rate;
}")
gg_dag(causal_diagram)
```

## Model Description

$$
\text{Exchange Rate Difference Scaled} \sim \text{Normal}(\mu_i, \sigma) 
$$
$$\mu_i \sim \beta_0 + \beta_1 * \text{Interest Rate Difference Scaled} + \beta_2 * \text{Inflation Rate Difference Scaled} + \beta_3 * \text{Growth Rate Difference Scaled} $$
$$
\beta_0 \sim \text{Normal}(0,1)
$$
$$
\beta_1 \sim \text{Normal}(0.5,1)
$$
$$
\beta_2 \sim \text{Normal}(-0.5,1)
$$
$$
\beta_3 \sim \text{Normal}(0.5,1)
$$
$$
\sigma \sim \text{Exponential}(1)
$$

## Rationals for Prior

Since I standardized all predictors and response variable, and are unsure of the relationship between those predictors and the response variable (there are ambivalent information about the direction of the affects of these predictors on exchange rate), so I have priors of the intercept and slope centers at 0 with standard deviation of 1. These priors suggests that I suspect if all predictors are kept at 0, the response variable would be at its mean (z-score = 0). And for all predictors, their relationship with the response variable are equally likely to be positive or negative, and most likely fall within -3 and 3 standard deviation. 



Higher Interest Rate, on the other hand, can attract foreign investers and drive up the currency value(https://www.investopedia.com/trading/factors-influence-exchange-rates/). Therefore, if dif_interest_rate (US - CN) increases, delta_dif_exchange_rate will increase, meaning US currency will worth more in China currency. For its slope, beta 1, I'm using a normal distribution center at 0.5 and has the standard deviation of 1. This means that for one standard deviation increase in differential in interest rate, the delta differential in exchange rate will most likely increase by 0.5 standard deviation. However, there is still possibility for a wider range of slope from negative to positive.

Higher inflation rate can depreciate the currency and thus lowers the exchange rate.  (https://www.investopedia.com/trading/factors-influence-exchange-rates/) . Therefore, if Dif_inflation_rate increases, delta_dif_exchange_rate should decrease. I'm using a normal distribution center at -0.5 and has the standard deviation of 1. This means that for one standard deviation increase in differential in inflation rate, I suspect the differential in delta exchange rate will most likely decrease by 0.5 standard deviation. However, there's still possibility for a wide range of positive and negative slopes.  


GDP measures growth. The better the growth, the better the currency for it attracts foreign investers, driving up the needs for the currency (https://www.economicshelp.org/macroeconomics/exchangerate/factors-influencing/). Therefore, if Dif_GDP increase,  delta_dif_exchange_rate should also increase. For the slope for Dif_GDP, Beta 3, I'm using a normal distribution centering at 0.5 with standard deviation of 1. This prior suggests a likely possitive relationship between GDP and exchange rate, but still keep the possibility for a wide range of both positive and negative values.

I suspect if all the above are kept at 0, there would be no change in exchange rate. Therefore I used a normal distribution that centers at 0 with standard deviation of 1 for Beta 0, the intercept. 

I'm using exponential(1) as the prior for sigma, with lower bound at 0, because the standard deviation cannot be 0. 

## Prior Predictive Distribution
```{r}
n_sim <- 500


predictors <- scaled_data |>
  select(Dif_Interest_Rate, Dif_Inflation_Rate, Dif_GDP) |>
  rename(interest = Dif_Interest_Rate,
         inflation = Dif_Inflation_Rate,
         growth = Dif_GDP) |>
  mutate(obs_id = row_number())

priors <- tibble(
  sim_id = 1:n_sim,
  beta_0 = rnorm(n_sim, 0, 1),
  beta_1 = rnorm(n_sim, 0.5, 1),
  beta_2 = rnorm(n_sim, -0.5, 1),
  beta_3 = rnorm(n_sim, 0.5, 1),
  sigma = rexp(n_sim, 1)
)

prior_pred <- crossing(priors, predictors) |>
  mutate(
    mu = beta_0 + beta_1 * interest + beta_2 * inflation + beta_3 * growth,
    y_sim = rnorm(n(), mean = mu, sd = sigma)
  )|>
  ungroup()

```
```{r}
gf_histogram(~y_sim,
             data = prior_pred)
```

The prior predictive distribution show the simulated delta_exchange_rate differences centers at 0, and have a range from about -12 to 12, which is large for a scaled z-score variable, but this wide range can make sure we don't exclude possible values. 



## Fitting Model


```{r}
stan_data <- compose_data(scaled_data)
```


```{r}

stan_program <- '
data {
  int<lower=1> n;     // number of observations
  vector[n] delta_exch_rate;    // response
  vector[n] Dif_Interest_Rate;     // predictor 1
  vector[n] Dif_Inflation_Rate;     // predictor 2
  vector[n] Dif_GDP;     // predictor 3
}
parameters {
  real<lower=0> sigma; // std of response, single continuous value
  real b0;
  real b1;
  real b2;
  real b3;
}
model {
  vector[n] mu; // vector of n values: expected 
  for (i in 1:n) { // loop over the n cases in the dataset to estimate mu_i values
    mu[i] = b0 + b1 * Dif_Interest_Rate[i] + b2 * Dif_Inflation_Rate[i] + b3 * Dif_GDP[i]; 
  }
  b0 ~ normal(0, 1); // prior for intercept
  b1 ~ normal(0.5, 1); // prior for both slope 1 (Interest)
  b2 ~ normal(-0.5, 1); // prior for both slope 2 (Inflation)
  b3 ~ normal(0.5, 1); // prior for both slope 3 (GDP)
  sigma ~ exponential(1); // prior for sigma
  delta_exch_rate ~ normal(mu, sigma); // defining likelihood in terms of mus and sigma
}
'

# fit the model
exch_rate_model <- rstan::stan(model_code = stan_program, data = stan_data)
exch_rate_model
```





